Finiteness in Cubical Type Theory

### 2020-06-01 sections/finiteness-predicates.tex
agda code works better
fully margined listings

  The support list for is \(\left[  \right]\), \(\left[ \text{tt} \right]\), and
  \(\left[ \text{false}, \text{true} \right]\), respectively, and the cover
  proof is a function that returns the index of the supplied element.
\end{proof}
\begin{marginfigure}%
  \begin{dottedbox}%
    % The code in our formalisation of these three proofs is quite simple.
    % Just as an example, the proof that \(\mathbf{Bool}\) is split enumerable is as
    % follows:
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  \end{dottedbox}
\end{marginfigure}
With the most basic simple types out of the way, the obvious next choice is the
(non-dependent) sums and products: \(\uplus\) and \(\times\).
Both of these types can be constructed from the \emph{dependent} sum, however,
so that is the type we will prove finite.
From that we can derive a much wider array of finiteness proofs.

@@

  Our task is to construct a proof of type:
  \begin{equation}
    \mathcal{E}!(\Sigma(x : A) , U(x))
  \end{equation}
  This proof itself is composed of two components:
  \begin{align}
    \mathit{support} &: \mathbf{List}(\Sigma(x : A) , U(x)) \\
    \mathit{cover} &: \Pi(x : \Sigma(y : A) , U(y)) , x \in \mathit{support}
  \end{align}

  To construct the support list, we apply the function \(\mathcal{E}!_U\) to
  every element in the support list of \(\mathcal{E}!_A\), extract the support
  lists from the resulting finiteness proofs, and concatenate them.

@@

\begin{agdalisting}
  This pattern of applying a function to each element in a list and
  concatenating the result is of course well-known in functional programming,
  and is in fact the pattern that makes lists a monad.
  While this insight isn't strictly relevant to our work here, it does mean
  the implementation of this function can use Agda's do notation, resulting
  in the following extremely clean implementation:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sup-sigma}
\end{agdalisting}
\subsection{Derivations}
We have a way to construct finiteness proofs, and a semiring-like toolbox to
combine them.
What we're now interested in is what we can \emph{derive} from them.

@@

\begin{agdalisting} \label{bool-slop}
  Consider the following two proofs of the finiteness of Bool:
  \begin{multicols}{2}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-slop}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  \end{multicols}
  While they both represent the ``same'' information (i.e. they both tell us
  that bool is finite), they are nonetheless not definitionally equal.
\end{agdalisting}

@@

  We can prove this goal by showing that \(\mathbf{Tuple}(n, U)\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \(\top\), which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}
\section{Cardinal Finiteness}
While we have removed some of the unnecessary information from our finiteness
predicates, one piece still remains.
\begin{agdalisting}
  The two following proofs are both valid proofs of the finiteness of
  \(\mathbf{Bool}\), and both do not include any duplicates.
  However they still differ:
  \begin{multicols}{2}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-rev}
  \end{multicols}
\end{agdalisting}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

>>>>

  The support list for is \(\left[  \right]\), \(\left[ \text{tt} \right]\), and
  \(\left[ \text{false}, \text{true} \right]\), respectively, and the cover
  proof is a function that returns the index of the supplied element.
\end{proof}
\begin{agdalisting}
  The code in our formalisation of these three proofs is quite simple.
  Just as an example, the proof that \(\mathbf{Bool}\) is split enumerable is as
  follows:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
\end{agdalisting}
With the most basic simple types out of the way, the obvious next choice is the
(non-dependent) sums and products: \(\uplus\) and \(\times\).
Both of these types can be constructed from the \emph{dependent} sum, however,
so that is the type we will prove finite.
From that we can derive a much wider array of finiteness proofs.

@@

  Our task is to construct a proof of type:
  \begin{equation}
    \mathcal{E}!(\Sigma(x : A) , U(x))
  \end{equation}
  This proof itself is composed of two components:
  \begin{align}
    \mathit{support} &: \mathbf{List}(\Sigma(x : A) , U(x)) \\
    \mathit{cover} &: \Pi(x : \Sigma(y : A) , U(y)) , x \in \mathit{support}
  \end{align}
  To construct the support list, we apply the function \(\mathcal{E}!_U\) to
  every element in the support list of \(\mathcal{E}!_A\), extract the support
  lists from the resulting finiteness proofs, and concatenate them.

@@

\begin{agdalisting}
  This pattern of applying a function to each element in a list and
  concatenating the result is of course well-known in functional programming,
  and is in fact the pattern that makes lists a monad.
  While this insight isn't strictly relevant to our work here, it does mean
  the implementation of this function can use Agda's do notation, resulting
  in the following extremely clean implementation:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sup-sigma}
\end{agdalisting}

\subsection{Derivations}
We have a way to construct finiteness proofs, and a semiring-like toolbox to
combine them.
What we're now interested in is what we can \emph{derive} from them.

@@

\begin{agdalisting} \label{bool-slop}
  Consider the following two proofs of the finiteness of Bool:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-slop}
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  While they both represent the ``same'' information (i.e. they both tell us
  that bool is finite), they are nonetheless not definitionally equal.
\end{agdalisting}

@@

  We can prove this goal by showing that \(\mathbf{Tuple}(n, U)\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \(\top\), which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}
\section{Cardinal Finiteness}
While we have removed some of the unnecessary information from our finiteness
predicates, one piece still remains.
\begin{agdalisting}
  The two following proofs are both valid proofs of the finiteness of
  \(\mathbf{Bool}\), and both do not include any duplicates.
  However they still differ:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-rev}
\end{agdalisting}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.
